Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/examples.tex
more terminology shifts

fun qs1 (l : list, k : unit) = ((), 
  case l of
    Empty => Leaf
  | Cons (h,t) => 
      Branch (
        let (left,right,n) = partition (h,t) in
        (n, #2 (qs1 left k), h, #2 (qs1 right k))
      )
  )`

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare (k, n) of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except that the mixed-stage argument to both passes is
is explicitly masked, and we do not suppress the trivial immediate result of
\texttt{qs1}.

>>>>

fun qs1 (l : list, k : unit) = 
  ((), case l of
    Empty => Leaf
  | Cons (h,t) => 
      Branch (
        let (left,right,n) = partition (h,t) in
        (n, #2 (qs1 left k), h, #2 (qs1 right k))))`

2`fun qs2 (((), k : int), p : tree) = 
  case p of
    Leaf => 0
  | Branch (n,left,h,right) =>
      case compare (k, n) of 
        LT => qs2 (((), k), left) 
      | EQ => h 
      | GT => qs2 (((), k-n-1), right)`
\end{lstlisting}
%
This is nearly identical to the cleaned-up code we presented in
\ref{fig:qs-split}, except that the mixed-stage argument to both passes is
is explicitly masked, and we do not suppress the trivial first-stage result
returned by \texttt{qs1}.

### 2015-02-28 paper/overview.tex
more terminology shifts

Suppose that we wish to compute a series of order statistics queries
on a list \texttt{l}. To this end, we can use
quickselect~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
function \texttt{qs} partitions the list by using the first element as
a pivot and then recurs on one of the two resulting sides to find the
desired element based on the relationship of \texttt{k} to the size of
the first half \texttt{i}.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permutation of the
input), function \texttt{qs} runs in expected linear time.
%
Using function \texttt{qs}, we can perform many order statistics
queries, for example with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
%
\begin{lstlisting}
qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time, making it
expensive to perform order statistics on more than a small number of
times. This is unfortunate: we wish to be able to map \texttt{qs} to
over another list and do so efficiently.

@@

Such transformations, called pass-separation by J{\o}rring and
Scherlis are commonly employed.  For example, as briefly mentioned in
\secref{intro}, modern graphics software is written exactly in this
way.  Unfortunately, as the example illustrates, multipass programs
are naturally complex.  We thus ask: is it would be possible to
perform such transformations automatically.  We answer this question
affirmatively.  Our approach is to start with a staged language and
then use a splitting algorithm to generate multi-pass programs from
staged programs.

@@

The idea behind staged programming is to use staging annotations or
types to indicate the stage of each expression.  To apply this
technique to our example, suppose that we have a language with two
worlds representing the two stages and a third, {\em ground} world
that contains terms with no staging.

>>>>

Suppose that we wish to compute a series of order statistics queries
on a list \texttt{l}. To this end, we can use
quickselect~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
function \texttt{qs} partitions the list by using the first element as
a pivot and then recurs on one of the two resulting sides to find the
desired element based on the relationship of \texttt{k} to the size of
the first half \texttt{i}.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permutation of the
input), function \texttt{qs} runs in expected $O(n)$ time.
%
Using function \texttt{qs}, we can perform many order statistics
queries, for example with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
%
\begin{lstlisting}
qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time, making it
expensive to perform order statistics on more than a small number of
times. This is unfortunate: we wish to be able to map \texttt{qs} to
over another list and do so efficiently.

@@

Such transformations, called pass-separation by J{\o}rring and
Scherlis are commonly employed.  For example, as briefly mentioned in
\ref{sec:intro}, modern graphics software is written exactly in this
way.  Unfortunately, as the example illustrates, multipass programs
are naturally complex.  We thus ask: is it possible to
perform such transformations automatically.  We answer this question
affirmatively.  Our approach is to start with a staged language and
then use a splitting algorithm to generate multi-pass programs from
staged programs.

@@

The idea behind staged programming is to use staging annotations or
types to indicate the stage of each subterm.  To apply this
technique to our example, suppose that we have a language with two
worlds representing the two stages and a third, {\em ground} world
that contains terms with no staging.

### 2015-02-28 paper/semantics.tex
large pass for consistency/small bugfixes
hyphenating "first-stage" and "second-stage" when appropriate
more terminology shifts

Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
\texttt{partition} it. The \texttt{case} expression in \texttt{qSelect} depends
on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
second-stage computations. Since \texttt{qss} itself produces a
$\fut\rmint$, the remainder of the function is second-stage code.

The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
expression) as a first-stage expression of type $\fut\rmint$ (the result 
of \texttt{qss}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
to \texttt{qss}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
second-stage $\rmint$.

@@

We evaluate \lang\ terms in two stages. The first stage of evaluation has three
mutually recursive modes of operation, depending on the world of the term it is
evaluating. Specifically, first stage evaluation:
\begin{itemize}
\item fully reduces terms in \bbonep, as they are purely first-stage;
\item reduces terms in \bbonem, save for any world \bbtwo\ terms they may
contain (inside \texttt{next}s);
\item scans terms in \bbtwo\ for any world \bbonem\ subterms (inside \texttt{prev}s), 
and reduces those in place.
\end{itemize}
These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
big-step judgments, respectively (defined in \ref{fig:diaSemantics}).

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val yhat = ...`
2`    val qs = yhat`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.

>>>>

Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
\texttt{partition} it. The \texttt{case} term in \texttt{qSelect} depends
on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
second-stage computations. Since \texttt{qss} itself produces a
$\fut\rmint$, the remainder of the function is second-stage code.

The $\next$ wraps a second-stage term of type $\rmint$ (the \texttt{case}
term) as a first-stage term of type $\fut\rmint$ (the result 
of \texttt{qss}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
to \texttt{qss}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
second-stage $\rmint$.

@@

We evaluate \lang\ terms in two stages. The first stage of evaluation has three
mutually recursive modes of operation, depending on the world of the term it is
evaluating. Specifically, first-stage evaluation:
\begin{itemize}
\item fully reduces terms in \bbonep, as they are purely first-stage;
\item reduces terms in \bbonem, save for any world \bbtwo\ terms they may
contain (inside \texttt{next}s);
\item scans terms in \bbtwo\ for any world \bbonem\ subterms (inside \texttt{prev}s), 
and reduces those in place.
\end{itemize}
These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
big-step judgments, respectively (defined in \ref{fig:diaSemantics}).

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`gr{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val yhat = ...`
2`    val qs = yhat`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.

### 2015-02-28 paper/splitting.tex
more terminology shifts

\paragraph {Unit.} 
As a simple first case, consider splitting the unit value, $\tup{}$
which masks to $\mval {\tup{}} {\tup{}}$.
%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
Splitting must produce a $c$ and $l.r$ satisfying,
\[
	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
Splitting $\pure{e}$ follows in similar fashion, except the immediate
result now becomes the result of evaluating the \bbonep\ term $e$.

@@

\paragraph {Function} 
Function introduction has a $\tup{}$ boundary value,
since functions are already fully reduced in our semantics.
However, since the body of a function may itself be multistage, splitting must continue into it.
The immediate result is a new function formed from the first-stage part of the original body.
The resumer is a new function formed out of the second-stage part of the original body.
It is the responsibility of the application site to save the precomputation of the function body
and pass it to the resumer version of the function.

>>>>

\paragraph {Unit.} 
As a simple first case, consider splitting the unit value, $\tup{}$
which masks to $\mval {\tup{}} {\tup{}}$.
%Units trivial contain unit split into more units: $\tup{} \vsplito \mval {\tup{}} {\tup{}}$.
Splitting must produce a $c$ and $l.r$ satisfying,
\[
	\reduce c {\tup{\tup{},b}} \text{ and } [b/l]r \equiv \tup{}
\]
Our algorithm uses $c=\tup{\tup{},\tup{}}$ and $l.r = \_.\tup{}$.
Splitting $\pure{e}$ follows in similar fashion, except the first 
component of $c$ now becomes the result of evaluating the \bbonep\ term $e$.

@@

\paragraph {Function} 
Function introduction has a $\tup{}$ boundary value,
since functions are already fully reduced in our semantics.
However, since the body of a function may itself be multistage, splitting must continue into it.
The $i$ is a new function formed from the first-stage part of the original body.
The resumer is a new function formed out of the second-stage part of the original body.
It is the responsibility of the application site to save the precomputation of the function body
and pass it to the resumer version of the function.
